Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2024-10-16 paper/main.tex
some edits to intro

\title{Bayesian Deep Gaussian Processes for Correlated Functional Data: 
        A Case Study in Cosmological Power Spectra}
\author{Stephen A. Walsh\thanks{Corresponding author: Division of Natural Sciences, 
        Math and Technology, Elms College, {\tt walshst@elms.edu}} \and 
        Annie S. Booth\thanks{Department of Statistics, North Carolina State University} \and
        David Higdon\thanks{Department of Statistics, Virginia Tech} \and
        Marco A.R. Ferreira\footnotemark[3] \and
        Kelly Moran\thanks{Los Alamos National Laboratory} \and
        Katrin Heitmann\thanks{Argonne National Laboratory}}
\date{\today}

@@

To further our understanding of the structure and movement of the universe, researchers utilize numerous cosmological surveys such as the Sloan Digital Sky Survey \citep{york2000sloan} and the upcoming Nancy Grace Roman Space Telescope \citep{Dore2019WFIRST} which are continuously growing in complexity. With these tools, cosmologists aim to continue learning more about cosmic acceleration \citep{caldwell2009physics}. To this end, one important ingredient to aid in this understanding of our universe is the matter power spectrum.

The matter power spectrum is a fundamental concept in cosmology and describes the distribution of matter as a function of spatial scale. It is often represented as a function of wavenumber $k$ (units Mpc$^{-1}$), which is inversely related to spatial scale. On large scales (i.e., small $k$ values), cosmic expansion dynamics and the matter power spectrum behave according to linear perturbation theory \citep{pietroni2008flowing, lesgourgues2009non}. On smaller scales, non-linear dynamics require the use of computationally intensive simulations; the Coyote Universe \citep{lawrence2010coyote} and more recently the Mira-Titan Universe \citep{moran2023mira} are two simulation suites dedicated to this effort.

>>>>

\title{Bayesian Deep Gaussian Processes for Correlated Functional Data: \\
        A Case Study in Cosmological Power Spectra}
\author{Stephen A. Walsh\thanks{Corresponding author: Division of Natural Sciences, 
        Math and Technology, Elms College, {\tt walshst@elms.edu}} \and 
        Annie S. Booth\thanks{Department of Statistics, North Carolina State University} \and
        David Higdon\thanks{Department of Statistics, Virginia Tech} \and
        Marco A.R. Ferreira\footnotemark[3] \and
        Kelly Moran\thanks{Los Alamos National Laboratory} \and
        Katrin Heitmann\thanks{Argonne National Laboratory}}
\date{\today}

@@

%To further our understanding of the structure and movement of the universe, 
%researchers utilize numerous cosmological surveys such as the Sloan Digital Sky 
%Survey \citep{york2000sloan} and the upcoming Nancy Grace Roman Space Telescope 
%\citep{Dore2019WFIRST} which are continuously growing in complexity. With these 
%tools, cosmologists aim to continue learning more about cosmic acceleration 
%\citep{caldwell2009physics}. To this end, one important ingredient to aid in this 
%understanding of our universe is the matter power spectrum. 
% @STEVE - I don't think these "surveys" are particularly relevant

Computer simulation experiments are invaluable tools in the study of cosmology.
Experiments which simulate the growth and expansion of the universe are growing
in prevalance and complexity \citep[e.g.,][]{lawrence2010coyote,moran2023mira}
\textcolor{red}{(more cites here?)}.  
We are particularly motivated by a simulation
of the matter power spectrum, which describes the distribution of matter as a 
function of spatial scale. 
%It is often represented as a function of wavenumber $k$ (units Mpc$^{-1}$), which is 
%inversely related to spatial scale. 
On large scales, power spectrum behave according to linear perturbation 
theory \citep{pietroni2008flowing, lesgourgues2009non}.  On smaller scales, nonlinear 
dynamics require the use of computationally intensive simulations, yielding
correlated functional data.
%the Coyote Universe \citep{lawrence2010coyote} and the Mira-Titan Universe \citep{moran2023mira} 
%are two simulation suites dedicated to this effort.  
From the highest level, our aim is to leverage such simulation data to 
first estimate and then predict underlying power matter spectra for various 
cosmological parameterizations.

%The expense of these simulations necessitates statistical ``emulators'' or
%``surrogate models'' \citep{santner2003design,gramacy2020surrogates}.
